Windowed least-squares model reduction for dynamical systems

### 2019-10-18 section4/spatial_numerical.tex
update

Defining a ``vectorization" function 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
the vectorized generalized coordinates over the $n$th time window are 
defined as \KTC{notationally, why are we doing a double bar instead of a bar?}
\EP{The notation for space--time in 2.4 uses bars. I don't want to confuse these generalized coordinates with those, they conceptually two different things} 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,1},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
We now define the weighted space--time residual over the entire window as
\KTC{To make this valid for seeping back into the previous time window, the
second argument could be omitted here (okay because there is already a
superscript on the residual) and arbitrary previous solutions could be
included in the first residual here.}
\begin{equation*}
\residLMSSlabArg{n} : \genstatecollocMatySlabArg{} \mapsto \begin{bmatrix}
 \sqrt{\frac{\quadWeightsLMSScalarArg{n}{1}}{2}} \stweightingMatOneArg{n} \residLMSArg{n,1}( \basisspaceArg{n} \genstateyDiscreteArgnt{1} + \stateInterceptArg{n}; 
%\residLMSArg{n,0}( \basisspaceArg{n} \genstateyDiscreteArgnt{0} + \stateInterceptArg{n}),
 \approxstateDiscreteArg{\indexMapper(n,0)},
\ldots , \approxstateDiscreteArg{\indexMapper(n,1 - \lmsWidthArg{n}{1}) }   ) \\
\vdots \\
 \sqrt{\frac{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}}}{2}} \stweightingMatOneArg{n} \residLMSArg{n,\nstepsArg{n}}( \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}} + \stateInterceptArg{n}; \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}-1} + \stateInterceptArg{n} ,  \ldots, \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n} - \lmsWidthArg{n}{\nstepsArg{n}} } + \stateInterceptArg{n} ) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{} \defeq \unroll (\genstateyDiscreteArg{1},\ldots,\genstateyDiscreteArg{\nstepsArg{n}} )$ and, for $i \le 0$,
\begin{equation}\label{eq:wls_genspatial_direct_args}
\genstateyDiscreteArgnt{n,i}  \equiv 
\begin{cases}
\genstateDiscreteArg{\indexMapper(n,i)}  & n = 2,\ldots,\nslabs,  \\
\genstateICOne  & n = 1. 
\end{cases}
\end{equation}
It is worth noting that, by design, 
\begin{equation*}
\objectiveArgLMS{n}_{\text{D-S}} \bigg( \basisspaceArg{n} \genstateDiscreteArg{n,1} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateDiscreteArg{n,\nstepsArg{n}} + \stateInterceptArg{n} \ \bigg) 
=
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard Gauss--Newton method. The algorithm consists of three fundamental steps: (1) compute the FOM O$\Delta$E residual given the current guess, (2) compute the Jacobian of the residual over the time window and (3) solve the linear least-squares problem and update the guess.

The practical implementation of the Gauss--Newton algorithm requires an efficient method for computing the Jacobian of the residual over the time window. In particular, it is important to note that for linear multistep methods (and many other time marching schemes) the Jacobian of the residual over the time window will yield an almost block-diagonal 
sparse matrix with the sparsity pattern
\begin{equation*}
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n})  =
 \begin{bmatrix*}[l]
\matshapeb & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*}.
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
This sparsity pattern can be leveraged to assemble the Jacobian of the residual over the time window from Jacobians of the residual at single time instances. 
Further, for solution techniques based on the normal equations, one can leverage the sparsity pattern 
to efficiently compute the Jacobian matrix--matrix product.
It is also worth noting that the 
normal equations will comprise a banded block system, this can be leveraged to speed up the linear solves at each Gauss--Newton iteration.

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces an extra dimension for parallelization that can be used to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$\genstatecollocMatSlabArg{n}_0 \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
$k \leftarrow 0$ \Comment{Set counter}\\
\While{\text{not converged}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
% Compute  $\Delta  \genstatecollocMatSlabArg{n} $ satisfying $ [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}

>>>>

Defining a ``vectorization" function 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
the vectorized generalized coordinates over the $n$th time window are 
defined as 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,1},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
We now define the weighted space--time residual over the entire window as
\begin{equation*}
\residLMSSlabArg{n} : \genstatecollocMatySlabArg{} \mapsto \begin{bmatrix}
 \sqrt{\frac{\quadWeightsLMSScalarArg{n}{1}}{2}} \stweightingMatOneArg{n} \residLMSArg{n,1}( \basisspaceArg{n} \genstateyDiscreteArgnt{1} + \stateInterceptArg{n}; 
%\residLMSArg{n,0}( \basisspaceArg{n} \genstateyDiscreteArgnt{0} + \stateInterceptArg{n}),
 \approxstateDiscreteArg{\indexMapper(n,0)},
\ldots , \approxstateDiscreteArg{\indexMapper(n,1 - \lmsWidthArg{n}{1}) }   ) \\
\vdots \\
 \sqrt{\frac{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}}}{2}} \stweightingMatOneArg{n} \residLMSArg{n,\nstepsArg{n}}( \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}} + \stateInterceptArg{n}; \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}-1} + \stateInterceptArg{n} ,  \ldots, \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n} - \lmsWidthArg{n}{\nstepsArg{n}} } + \stateInterceptArg{n} ) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{} \equiv \unroll (\genstateyDiscreteArg{1},\ldots,\genstateyDiscreteArg{\nstepsArg{n}} )$ and, for $i \le 0$,
\begin{equation}\label{eq:wls_genspatial_direct_args}
\genstateyDiscreteArgnt{n,i}  \equiv 
\begin{cases}
\genstateDiscreteArg{\indexMapper(n,i)}  & n = 2,\ldots,\nslabs,  \\
\genstateICOne  & n = 1. 
\end{cases}
\end{equation}
We note that
\begin{equation*}
\objectiveArgLMS{n}_{\text{D-S}} \bigg( \basisspaceArg{n} \genstateDiscreteArg{n,1} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateDiscreteArg{n,\nstepsArg{n}} + \stateInterceptArg{n} \ \bigg) 
=
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard
Gauss--Newton method. Each Gauss--Newton iteration consists of three fundamental steps: (1)
compute the FOM O$\Delta$E residual given the current guess, (2) compute the
Jacobian of the residual over the time window, and (3) solve the linear least-squares problem and update the guess.

The practical implementation of the Gauss--Newton algorithm requires an
efficient method for computing the Jacobian of the residual over the time
window
$\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}$.
For this purpose, we can leverage the fact that this Jacobian is a block lower
triangular matrix with the following sparsity pattern (for $\lmsWidthArg{n}{i} = 1$,
$i=1,\ldots,\nstepsArg{n}$):  \KTC{make sure the column widths are correct.
Maybe add dimensions for each block?}
\begin{equation*} \begin{bmatrix*}[l]
\matshapeb & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*}.
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
In particular, solution techniques can leverage this structure, e.g., to
efficiently compute Jacobian--vector products.
Another consequence of this sparsity pattern is that the normal equations
arising at each Gauss--Newton iteration 
comprise a banded block system that can also be exploited.

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the
	Gauss--Newton Method)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton--Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces
	an extra dimension for parallelization that can be exploited to furture
	reduce the wall time. The investigation of these additional, potentially
	more efficient, solution algorithms is a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$\genstatecollocMatSlabArg{n}_0 \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
$k \leftarrow 0$ \Comment{Set counter}\\
\While{\text{not converged}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
% Compute  $\Delta  \genstatecollocMatSlabArg{n} $ satisfying $ [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}
